---
title: "ESA 2021"
output:
  learnr::tutorial:
    css: css/custom.css
    fig_width: 5.5
    fig_height: 5.5
    # ace_theme: "cobalt"
    progressive: false
    toc: true
    toc_depth: 6
    toc_float:
      collapsed: false
      smooth_scroll: false
runtime: shiny_prerendered
description: "Community analysis of species, traits, phylogenies, and responses to environment."
---

```{r begin, include=FALSE, eval=TRUE}
require(learnr)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, include = TRUE, out.width="99%")
knitr::knit_theme$set("rootwater")
learnr::tutorial_options(exercise.timelimit = 10, exercise.eval = FALSE, exercise.completion = TRUE)
```

```{r setup, include=FALSE}
### HIDDEN ###
require(vegan)
require(labdsv)
require(ade4)
require(ecodist)
require(fso)
require(vegclust)
require(ape)
require(picante)
require(mgcv)
load('veg.rda')  # assumed to be in your working dir
xy  <- veg$xy    # spatial
spe <- veg$spe   # species
env <- veg$env   # environment
tra <- veg$tra   # traits
phy <- veg$phy   # phylogeny
rm(veg)          # cleanup
spe <- data.frame(log10(spe + 1))
env <- data.frame(decostand(scale(env, center=F), 'range'))
tra <- data.frame(decostand(tra, 'range'))
d   <- vegdist(spe, method='bray', binary=T)
D   <- stepacross(d, 'shortest', toolong=1, trace=F)
E   <- dist(xy)                         # spatial distance matrix
pc  <- pcnm(E, w=rowSums(spe)/sum(spe)) # PCNMs *weighted*  by abundances
cl  <- vegclustdist(D, mobileMemb=7, method='FCM', m=1.2, nstart=5)
grp <- defuzzify(cl, 'max')[[2]]
m1  <- metaMDS(D, k=2, try=200, trymax=500, trace=0)  # NMS
m2  <- cmdscale(D, k=2, add=T)                        # PCoA
m3  <- prcomp(spe)                                    # PCA
grp_k <- cut(env$k, quantile(env$k, probs=seq(0,1,by=0.25)), include.lowest=T,
               labels=c('lo','med','hi','veryhi'))
`get_palette` <- function() {
  pal <- c('#414487E6','#404688E6','#3F4889E6','#3E4989E6','#3E4C8AE6',
           '#3D4E8AE6','#3C508BE6','#3B528BE6','#3A548CE6','#39558CE6',
           '#38588CE6','#375A8CE6','#365C8DE6','#355E8DE6','#35608DE6',
           '#34618DE6','#33638DE6','#32658EE6','#31678EE6','#30698EE6',
           '#306A8EE6','#2F6C8EE6','#2E6E8EE6','#2D708EE6','#2C718EE6',
           '#2C738EE6','#2B748EE6','#2A768EE6','#2A788EE6','#297A8EE6',
           '#287C8EE6','#287D8EE6','#277F8EE6','#26818EE6','#26828EE6',
           '#25848EE6','#24868EE6','#24878EE6','#23898EE6','#228B8DE6',
           '#228D8DE6','#218F8DE6','#21908CE6','#20928CE6','#20938CE6',
           '#1F958BE6','#1F978BE6','#1F998AE6','#1F9A8AE6','#1E9C89E6',
           '#1F9E89E6','#1FA088E6','#1FA187E6','#20A386E6','#20A486E6',
           '#21A685E6','#22A884E6','#24AA83E6','#25AC82E6','#26AD81E6',
           '#28AE80E6','#2AB07FE6','#2DB27DE6','#2FB47CE6','#32B67AE6',
           '#34B679E6','#37B878E6','#3ABA76E6','#3DBC74E6','#40BD72E6',
           '#43BF71E6','#47C06FE6','#4AC16DE6','#4EC36BE6','#52C569E6',
           '#55C668E6','#59C864E6','#5DC863E6','#60CA60E6','#65CB5EE6',
           '#68CD5BE6','#6DCD59E6','#71CF57E6','#75D054E6','#7AD151E6',
           '#7FD34EE6','#83D44CE6','#87D549E6','#8CD646E6','#90D743E6',
           '#95D840E6','#9AD93CE6','#9FDA3AE6','#A3DA37E6','#A8DB34E6',
           '#ADDC30E6','#B2DD2DE6','#B7DE2AE6','#BBDF27E6')
  return(pal)
}
`colvec` <- function(x) {
  pal <- get_palette()
  return(pal[cut(as.numeric(x), breaks=length(pal), include.lowest=TRUE)])
}
`makecwm` <- function (spe, tra) {
  spe <- as.matrix(spe)
  tra <- as.matrix(tra)
  `standardize` <- function(x) {
    (x - min(x, na.rm=TRUE)) / diff(range(x, na.rm=TRUE))
  }
  tra <- apply(tra, MARGIN = 2, FUN = standardize)
  awt <- spe %*% tra
  awt / rowSums(spe, na.rm=TRUE)
}
cwm <- data.frame(makecwm(spe, tra)) # make the CWM traits matrix
### HIDDEN ###
```


<!-- --------------------------------------------- -->


## Overview

#### __*Why are we here today?*__ {#why-here}

This course demonstrates **the practice of ecological community analysis in R**. Our goal is to link biological reality with corresponding conceptual and mathematical models.  We borrow and apply concepts from ecology, statistics and computer science, but do not get too bogged down in theoretical details.  This short course aims to be an [axe handle](https://www.poetryfoundation.org/poems/57150/axe-handles){target="_blank"} --- carve and shape it to your own practice.


### Learning outcomes

We will: 

1.  Analyze, visualize and interpret ecological community data in R software.  
2.  Select, modify or create new multivariate methods for ecological community analysis in R software.  
3.  Gain proficiency with typical R software usage through direct experience. 


### Assumptions

We assume you have a basic familiarity with R.  If not, please peruse [here](http://ecology.msu.montana.edu/labdsv/R/labs/R_ecology.html).  At the least, you should be able to assign an object, distinguish between different objects (vector, matrix, data.frame, etc.), and index rows/columns of a matrix or data.frame.

We assume you have familiarity with basic statistics (population vs sample; mean, SD, SE, CI; simple linear models), and have at least dabbled in linear algebra (matrices; linear combinations), and perhaps become suspicious that calculus may indeed exist.  If not, you could brush up on [matrices](https://courses.lumenlearning.com/boundless-algebra/chapter/introduction-to-matrices/).

### Software installation

Packages contain supplemental functions you will need.  They are supplemental beyond "base" R.  Ordinarily you would open R and install required packages as follows (but these dependencies should have automatically installed when you first installed this tutorial). 

```{r install-pkgs, exercise=FALSE, eval=FALSE}
install.packages('vegan')     # community ecology
install.packages('labdsv')    # community ecology
install.packages('ade4')      # community ecology
install.packages('ecodist')   # community ecology
install.packages('fso')       # ordination
install.packages('vegclust')  # clustering
install.packages('ape')       # phylogenetics
install.packages('picante')   # phylogenetics
install.packages('mgcv')      # nonlinear regression
```

### Load packages

To use functions provided by packages, load them into your R environment.  Again, this has already been done for you in the tutorial.

```{r load-pkgs, exercise=FALSE, eval=FALSE}
require(vegan)
require(labdsv)
require(ade4)
require(ecodist)
require(fso)
require(vegclust)
require(ape)
require(picante)
require(mgcv)
```

### Useful functions to explore

Base R already contains lots of useful functions.  This list has a few basic functions: 

```
ls()
dim(spe)
NROW(spe)
NCOL(spe)
str(spe)
head(spe)
names(spe)
x <- spe[,11]
x <- spe[,'bolboschoenus_maritimus']
hist(x)
plot(x)
rm(x)
```
From the list above, copy-and-paste some commands into the code box below, and help describe a dataset called `spe`.  What is inside that object?  What size is it?

```{r blank-to-try-functions, exercise=TRUE}

### run code here

```

### Define objects and functions

We've already said that you can use functions from base R or from other packages.  But you can also "roll your own"!  Functions take the generic form `f <- function(x) { x }`.  For example:

```{r define-functions, exercise=TRUE}
# function to pick an item from the vector by position
`item_picker` <- function(x, pos = NULL) {
  x[pos]
}
z <- 1:19               # assign a sequence of numbers to object `z`
item_picker(z, pos = 5) # pick the fifth element of `z`
```

## Data

#### __*Where did this dataset originate?*__ {#dataorigin}

### Mafragh, Algeria vegetation data

We will use one core dataset throughout this tutorial. The `veg` dataset gives information about spatial coordinates, species, environment, traits, and phylogeny for plants on the Mafragh coastal plain in North Africa.  Specifically, `veg` is a list containing five items:

* `xy` 97 observations of 2 spatial coordinates
* `spe` 97 observations of 56 plant species
* `env` 97 observations of 11 soil environmental variables
* `tra` 12 traits for 56 plant species
* `phy` phylogeny for 56 plant species

### Load data

A common task in R is to read from CSV files, e.g., `x <- read.csv('path/to/file.csv')`, but here we will load data that's already in R format.

```{r data-load, exercise=FALSE, eval=FALSE}
load('veg.rda')  # assumed to be in your working dir
xy  <- veg$xy    # spatial
spe <- veg$spe   # species
env <- veg$env   # environment
tra <- veg$tra   # traits
phy <- veg$phy   # phylogeny
rm(veg)          # cleanup
ls()             # objects now in this environment
```

### Study area description

The Mafragh plain is in El Tarf Province of far northeast Algeria (36.84704$^{\circ}$, 7.945$^{\circ}$).  According to documentation in R package `ade4`:

> This marshy coastal plain is a geomorphological feature, east of Port of Annaba, limited to the north by the Mediterranean Sea and a dune cordon, to the south by clay-sandstone Numidian massifs, to the west by a wadi and to East by an irrigated perimeter. It covers 15,000 ha of which 10,000 form the area of study.

Potential human pressures include grazing, irrigation and salination, plowing, and housing development.  Its Köppen climate type is "Csa: Hot-summer Mediterranean climate".

### History of the Mafragh dataset

The data originate from the field studies of Prof. Gérard de Bélair of Annaba in northeast Algeria.  Born in France, de Bélair became a missionary priest in 1969 shortly after Algeria's independence, then an agronomist during the Agrarian Revolution, then a botanist, then professor, then conservationist (according to his [lay](http://gdebelair.com/itineraire.html) and [clergy](https://www.visages-diocese-autun.fr/visage/de-belair-gerard/) accounts).  As of 2021, he continues to actively publish on the vegetation of northern Algeria.  

We should recognize the colonialist legacies that brought us here.  France controlled Algeria from 1830 until its independence in 1962.  Prior to that, Algeria was occupied by Ottoman Turks, Spaniards, Arab Muslims, Berber/Amazigh, Romans, Phoenicians, and prehistoric occupants dating to at least 1.8 million years ago.  We may speculate that the region's vegetation and cultural histories have been closely intertwined.  

The data were included in early versions of the `ade4` software (Dray and Dufour 2007) which itself has its roots at Université Lyon 1, France.  The present data are lightly modified from Appendix S4 of Pavoine et al. (2011), who also added a phylogeny.

### Key references

de Bélair, G. 1981. Biogéographie et aménagement de la Plaine de la Mafragh (Annaba - Algérie) [Biogeography and development of the Mafragh Plain (Annaba, Algeria)]. Dissertation, Université Paul Valéry, Montpellier. http://www.sudoc.fr/00871262X  

de Bélair, G. 1990. Structure, fonctionnement et perspectives de gestion de quatre écocomplexes lacustres et marécageux (El Kala, Est-Algérien) [Structure, function and management perspectives of four lake and marsh eco-complexes (El Kala, East Algeria)]. Dissertation, Université des sciences et techniques de Montpellier 2. http://www.sudoc.fr/007729197  

de Bélair, G. and M. Bencheikh-Lehocine. 1987. Composition et déterminisme de la végétation d'une plaine côtière marécageuse: La Mafragh (Annaba, Algérie) [Composition and determinism of the vegetation of a marshy coastal plain: La Mafragh (Annaba, Algeria)]. Bulletin d'Ecologie, 18(4): 393–407.  

Dray, S. and A. Dufour. 2007. The `ade4` package: implementing the duality diagram for ecologists. Journal of Statistical Software 22(4): 1-20. [doi:10.18637/jss.v022.i04](https://doi.org/10.18637/jss.v022.i04)  

Pavoine, S., Vela, E., Gachet, S., de Bélair, G. and Bonsall, M. B. 2011. Linking patterns in phylogeny, traits, abiotic variables and space: a novel approach to linking environmental filtering and plant community assembly. Journal of Ecology 99: 165–175. [doi:10.1111/j.1365-2745.2010.01743.x](https://doi.org/10.1111/j.1365-2745.2010.01743.x)  


<!-- --------------------------------------------- -->


## Pre-analysis

#### __*How do I get my data into shape for analysis?*__ {#pre-analysis}

### Check data structure

A first step in exploring ecological data is to examine its structure.  Copy-and-paste some of this code into the box below, and run one at a time:

```
str(spe)     # community (species) abundance matrix
str(env)     # environmental matrix
str(tra)     # traits matrix
str(phy, 1)  # phylogeny
str(xy)      # spatial coordinates
```

```{r structure, exercise=TRUE}

### run code here

```

### Find and replace troublesome values

```{r replace-pesky, exercise=TRUE}
### create example matrix with some pesky values
s <- spe
s[5,2] <- NA
s[6,4] <- (-999)

### index the row and column of any NA values
which(is.na(s), arr.ind=TRUE)

### index the row and column of any negative values
which(s < 0, arr.ind=TRUE)

### replace value by logical test
s[is.na(s)] <- 777
head(s)
```

### Data transformations

```{r transformations, exercise=TRUE}
### load data from file
load('veg.rda')  # assumed to be in your working dir
spe <- veg$spe   # species
env <- veg$env   # environment
tra <- veg$tra   # traits

### basic transformations
spet <- data.frame(log10(spe + 1))
envt <- data.frame(vegan::decostand(scale(env, center=F), 'range'))
trat <- data.frame(vegan::decostand(tra, 'range'))

### compare
par(mfrow=c(1,2))
plot(tra$bfp, trat$bfp)
plot(env$k2o, envt$k2o)
```

### Abundances to presence/absence

Sometimes we simply need to convert abundance values (counts or percentages) to binary presence/absence values (0 or 1).

```{r presence-absence, exercise=TRUE}
s <- (spe > 0) * 1  # from numeric to 0/1
s[1:5, 1:7]         # peek at a few
range(s)            # is this what you expect?
```

### Outliers

A quick way to search for _multivariate_ outliers is to use the distance/dissimilarity matrix.  An "outlier" is the sample unit(s) which exceed a defined number of SDs away from the grand mean (i.e., a large _z_-score) (McCune and Grace 2002).

```{r outliers-find, exercise=TRUE}
### define outlier function
`outliers` <- function (x, mult=2, method='bray') {
  d <- as.matrix(vegan::vegdist(x, method=method, binary=F, diag=T, upper=T))
  diag(d) <- as.numeric(1)     # avoid zero-multiplication
  m       <- apply(d, 2, mean) # site means
  z       <- scale(m)          # z-scores
  data.frame(mean_dist = m, z = z, is_outlier = abs(z) >= mult)
}

### try it
o <- outliers(spe, mult=2)
head(o, 7)
which(o$is_outlier) 
### Thinker: how would you find the z-score of these outliers?
```

### Test validity of species matrix

Before proceeding with most multivariate analyses, our species abundance matrix must have 1) no missing values, 2) every site must have at least one species observation, and 3) every species must occur in at least one site.

```{r mx-valid, exercise=TRUE}
!anyNA(spe)                     # expect TRUE, no missing values
all(rowSums(spe, na.rm=T) != 0) # expect TRUE, no empty sites
all(colSums(spe, na.rm=T) != 0) # expect TRUE, no empty species
```

### Visualize data

Let's map the spatial coordinates.

```{r viz-xy, exercise=TRUE}
# spatial
plot(xy, pch=19, col='grey', xlab='Eastness', ylab='Northness')
```

Plot the species abundance matrix as a heatmap.

```{r viz-spe, exercise=TRUE}
# species
vegan::tabasco(spe, col=get_palette())
```

Plot the soils (environmental) matrix as a heatmap.

```{r viz-env, exercise=TRUE}
# environment
vegan::tabasco(env, col=get_palette())
```

Plot the traits matrix as a heatmap.

```{r viz-tra, exercise=TRUE}
# traits
vegan::tabasco(tra, col=get_palette())
```

Plot the phylogenetic tree.

```{r viz-phy, exercise=TRUE}
# phylogeny
plot(phy, cex=0.6, no.margin=TRUE)
```

### Key references

McCune, B., and J.B. Grace. 2002. _Analysis of Ecological Communities_. MjM Software Design, Gleneden Beach, Oregon, USA.  


<!-- --------------------------------------------- -->


## Diversity measures

#### __*How diverse is this site? many sites? region?*__ {#diversity-main}

Let's examine the "diversity" of our sites and study area.  Diversity can encompass any sort of variation, e.g., taxonomic, trait, or phylogenetic.  Here we assume that `spe` is a species abundance matrix where rows = sites, columns = taxa, and values must not be negative or missing.

### Gamma (regional) diversity

```{r div-gam, exercise=TRUE}
gamma <- sum(colSums(spe) > 0)
gamma
```

### Alpha (per-site) diversity

```{r div-alf, exercise=TRUE}
alpha <- rowSums(spe > 0)
alpha    # within-site
avgalpha <- mean(rowSums(spe > 0))
avgalpha # average within-site
```

### Beta (among-site) diversity: Whittaker's

```{r div-bet, exercise=TRUE}
gamma    <- sum(colSums(spe) > 0)
avgalpha <- mean(rowSums(spe > 0))
beta     <- gamma / avgalpha - 1
beta
```

### Beta diversity: dust bunny indices

McCune and Root (2015) proposed two model-free measures of how strongly multivariate species data exhibit a "dust bunny" distribution, that is, departure from multivariate normality because of many zero abundances due to few species occurring in any given sample unit.  Quantifying this is important because multivariate normality is a core assumption of many analytical tools (PCA, CCA, linear regression models).

```{r div-propz, exercise=TRUE}
### 1 -- proportion of zeros in the matrix
###   (independent of abundance)
eps      <- .Machine$double.eps # machine tolerance
propzero <- sum(spe < eps) / prod(dim(spe))
cat('Proportion of zeros in matrix:', propzero, '\n')

### 2 -- "dust bunny index" of McCune and Root (2015)
###   (integrates abundances)
dbi <- 1 - mean(as.matrix(vegan::decostand(spe, method='max')))
cat('Dust bunny index:', dbi, '\n')
```

### Beta-diversity: no-share sites

High species turnover (beta-diversity) can cause many site-pairs to share no species in common.  What proportion of all pairs of sites have totally different sets of species?

```{r div-noshare, exercise=TRUE}
### how many site-pairs share no species in common?
z <- vegan::no.shared(spe)
propnoshare <- sum(z) / length(z)
cat('Proportion of no-share sites:', propnoshare, '\n')
```


### Trait diversity

[Here.](#traits-main)


### Phylogenetic diversity

[Here.](#phylo-main)


### Key references

McCune, B., and H.T. Root. 2015. Origin of the dust bunny distribution in ecological community data. _Plant Ecology_ 216(5): 645-656.

<!-- --------------------------------------------- -->


## Dissimilarities

#### __*How dissimilar are community compositions among sites?*__ {#dissim-main}

Dissimilarities among sites help us position sites along gradients, organize them into clusters, or distinguish differences among groups of sites.  

### Dissimilarity matrix for species

Create a pairwise dissimilarity matrix `D` based on shared and unshared species among sites.  Euclidean distances are seldom useful for species abundance data, here we use the more preferable Bray-Curtis (Sørensen) dissimilarity.  Legendre and Legendre (2012) give a full survey of existing distance/dissimilarity measures.

```{r dissimilarities, exercise=TRUE}
d <- vegdist(spe, method='bray', binary=T)
str(d)
tabasco(as.matrix(d), col=get_palette()) 
### Thinker: why is the diagonal 'blank'?
```

### Stepacross adjustment

Sometimes a pair of sites are "maximally dissimilar": they share no species at all, so pairwise dissimilarity is impossible to quantify using a proportional measure (e.g., Bray-Curtis).  _Stepacross adjustment_ solves this by linking "no-share" sites with intermediate site(s) that share at least some species in common.  Smith (2017) describes other possible solutions.  

```{r stepacross, exercise=TRUE}
D <- stepacross(d, 'shortest', toolong = 1)
plot(d, D, xlab = 'Original', ylab = 'Stepacross')
### Thinker: why the vertical stripe near 1.0?
```

### Distance matrix for environment

Create a pairwise distance matrix `E` based on Euclidean environmental distances among sites.

```{r dist-env, exercise=TRUE}
E <- vegdist(env, method='bray', binary=T)
str(E)
tabasco(as.matrix(E), col=get_palette())
```

### Key references

Legendre, P., and L. Legendre. 2012. _Numerical Ecology_. Volume 24, Third Edition. Elsevier.  

Smith, R.J. 2017. Solutions for loss of information in high-beta-diversity community data. _Methods in Ecology and Evolution_ 8:68–74.  


<!-- --------------------------------------------- -->


## Ordination (unconstrained)

#### __*How best to order sites by dissimilarities of community compositions?*__ {#ord-unconstrained}

Order sites according to community compositions.  Ordination distances indicate compositional dissimilarities.  Nearby points are more compositionally similar than distant points.

```{r ordination, exercise=TRUE, exercise.timelimit=120}
# three kinds of ordination
m1 <- metaMDS(D, k=2, try=200, trymax=500)  # NMS
m2 <- cmdscale(D, k=2, add=T)               # PCoA
m3 <- prcomp(spe)                           # PCA
```

For ecological community data, NMS (Kruskal 1964) is almost always the best choice (Minchin 1987).


### Visualizing ordinations

```{r ordination-plotting-01, exercise=TRUE}
# color vector for plotting
u <- get_palette()
u <- u[1:nrow(spe)]

# compare three kinds of ordination
par(mfrow=c(1,3), bty='l', las=1)
plot(m1$points, type='n', xlab='NMDS1', ylab='NMDS2')
text(m1$points, rownames(m1$points),cex=.8, col=u)
plot(m2$points, type='n', xlab='PCoA1', ylab='PCoA2')
text(m2$points, rownames(m2$points),cex=.8, col=u)
plot(m3$x, type='n', xlab='PCA1', ylab='PCA2')
text(m3$x, rownames(m3$x),cex=.8, col=u)
par(mfrow=c(1,1))
```

Plotting again, now overlay environmental variables. 

```{r ordination-plotting-02, exercise=TRUE}
### overlay gradients on the NMS using point colors
par(mfrow=c(1,3), bty='l', las=1)
plot(scores(m1), pch=16, col=colvec(env$k))
plot(scores(m1), pch=16, col=colvec(env$k2o))
plot(scores(m1), pch=16, col=colvec(env$sand))
par(mfrow=c(1,1))

### overlay gradients on the NMS by fitting a GAM surface
f1 <- ordisurf(m1 ~ env$k, plot = FALSE)
plot(scores(m1), pch = 16, col = colvec(env$k))
plot(f1, add = TRUE, col = 1, lwd = 2)
```

### Ordination goodness-of-fit

Ordination methods have intrinsic measures of fit (e.g., stress for NMS).  However, for all methods, a "good" ordination should have a decent correlation between the Euclidean inter-point ordination distances versus the original species dissimilarities used to construct the ordination.

```{r ordination-goodness-of-fit, exercise=TRUE}
cor(dist(m1$points), D, method='kendall') # NMS almost always best
cor(dist(m2$points), D, method='kendall')
cor(dist(m3$x),      D, method='kendall')
### Thinker: why is this unfair to PCA?
```

We can also use Procrustes analysis (Peres-Neto and Jackson 2001) to examine the fit of a species ordination versus an environmental ordination.

```{r ord-fit-to-env, exercise=TRUE}
pc <- prcomp(env[,sapply(env, is.numeric)], scale=T) # environmental PCA
pc <- scores(pc, display='sites', choices=1:2)       # environmental PCA scores
eD <- vegdist(pc, method='euc')  # Euclidean interpoint distances
protest(m1$points, pc)           # fit = correlation in Procrustes rotation
protest(D, eD)                   # fit = correlation in Procrustes rotation
```

### Dimensionality selection in NMS

For NMS, the user must select the number of dimensions _a priori_.  Yet, what is the "best" number of dimensions?  We can examine how stress changes as we increment the number of dimensions, and select the dimensionality that balances sufficiently "low" stress with the fewest dimensions.  

```{r dimensionality-selection, exercise=TRUE}
# define screeplot function, running NMS for varying 'k' dimensions
`scree_nms` <- function(D, k=5, ...) {
  stress <- rep(NA, k)
  for (i in 1:k) {
    cat('calculating', i, 'of', k, 'dimensions...\n')
    stress[i] <- metaMDS(D, k=i, trace=0, ...)$stress
  }
  plot(1:k, stress, main='', xlab='Dimension', ylab='Stress',
       ylim=c(0, max(stress)*1.05), pch=16, las=1, bty='l')
  lines(1:k, stress)
  abline(0.20, 0, col='red', lty = 2)
  data.matrix(stress)
}
scree <- scree_nms(D, k=5, trymax=10)
scree
### Thinker: how many dimensions would you select?
```

### Key references

Kruskal, J.B. 1964. Multidimensional scaling by optimizing goodness-of-fit to a nonmetric hypothesis. _Psychometrika_ 29: 1–28.

Minchin, P.R. 1987. An evaluation of relative robustness of techniques for ecological ordinations. _Vegetatio_ 69: 89–107.  

Peres-Neto, P. R., and D. A. Jackson. 2001. How well do multivariate data sets match? The advantages of a Procrustean superimposition approach over the Mantel test. _Oecologia_ 129:169–178.  


<!-- --------------------------------------------- -->


## Ordination (constrained)

#### __*How best to order sites by dissimilarities of community compositions, in response to measured environmental variables?*__ {#ord-constrained}

Constrained ordination addresses specific _a priori_ hypotheses about variation in community compositions _attributable to measured environmental gradients_.  It portrays only a fragment of all community variation, and is not used for exploratory analysis.  

Two important tools are distance-based RDA (dbRDA, McArdle and Anderson 2001), and fuzzy set ordination (FSO, Roberts 2008), both of which let the user select their own dissimilarity measure.  Other methods (redundancy analysis or constrained correspondence analysis) force an implicit dissimilarity measure and make linearity/distributional assumptions.  

```{r ord-constrained, exercise=TRUE}
### RDA: redundancy analysis (assumes linear species responses)
m1 <- rda(spe ~ k + sand + conductivity, data=env)

### CCA: constrained correspondence analysis (assumes unimodal species responses)
m2 <- cca(spe ~ k + sand + conductivity, data=env)

### dbRDA: distance-based RDA
m3 <- dbrda(D ~ k + sand + conductivity, 
            data = env, 
            comm = spe, 
            add  = 'lingoes')

### FSO: fuzzy set ordination
m4 <- fso::fso( ~ k + sand + conductivity, 
                dis     = D, 
                data    = env, 
                permute = F)

### summaries
print(m1)
print(m2)
print(m3)
summary(m4)

### compare four kinds of ordination
u <- colvec(env$k)         # color by soil potassium
par(mfrow=c(2,2), bty='l', las=1)
plot(m1, disp='wa')
points(m1, pch=16, col=u)
plot(m2, disp='wa')
points(m2, pch=16, col=u)
plot(m3, disp='wa')
points(m3, pch=16, col=u)
plot(scores(m4$mu[,1:2]), pch=16, col=u, asp=1,
     xlab='MFSO1, potassium', ylab='MFSO2, sand')
```

### Key references

McArdle, B.H. & Anderson, M.J. 2001. Fitting multivariate models to community data: a comment on distance-based redundancy analysis. _Ecology_ 82: 290-297.  

Roberts, D.W. 2008. Statistical analysis of multidimensional fuzzy set ordinations. _Ecology_ 89: 1246-1260.



<!-- --------------------------------------------- -->


## Group clustering

#### __*How to group sites according to dissimilarity of community compositions?*__ {#clustering}


### Hierarchical: Ward's clustering

Ward's clustering is a hierarchical cluster analysis method based on community dissimilarities (Murtagh and Legendre 2014).

```{r wards, exercise=TRUE, fig.asp=0.50}
### perform the clustering
k   <- 7                           # number of groups is specified in advance
cl  <- hclust(D, method='ward.D2') # clustering solution
grp <- cutree(cl, k)               # group memberships

### plot the dendrogram, and plot the groups onto an NMS ordination
u <- colvec(grp)
par(mfrow=c(1,2), bty='l', las=1, oma=c(0,0,0,0), mar=c(3,3,0.5,0.5))
plot(cl, cex=0.5, main='')
rect.hclust(cl, k, border=unique(u))
plot(scores(m1), pch=16, cex=0.7, col=u)
ordicluster(m1, cl, prune=k, lwd=0.5, col=u)
```


### Non-hierarchical: fuzzy clustering

Non-hierarchical methods also exist.  And, it may be useful to define groups based on _fuzzy_ rather than _crisp_ partitions, using fuzzy clustering (de Caceres et al. 2010).

```{r fuzz-01, exercise=TRUE, exercise.timelimit = 20}
ns   <- 1       # number of random starts (increase for real analyses!)
k    <- 7       # number of groups is specified in advance
# fuzzy c-means
cl   <- vegclustdist(D, mobileMemb=k, method='FCM', m=1.2, nstart=ns)
# fuzzy c-means with a noise cluster
cl_a <- vegclustdist(D, mobileMemb=k, method='NC', m=1.2, dnoise=0.8, nstart=ns)
# fuzzy c-medoids
cl_b <- vegclustdist(D, mobileMemb=k, method='FCMdd', m=1.2, nstart=ns)
# crisp k-means
cl_c <- vegclustdist(D, mobileMemb=k, method='KM', nstart=ns)

# how well do the other methods agree with fuzzy c-means?
concordance(cl, cl_a)
concordance(cl, cl_b)
concordance(cl, cl_c)

```

Let's examine group memberships `grp` from the fuzzy clustering.  While memberships are fuzzy, we can ask for a _crisp_ membership based on a threshold or else the maximum fuzzy membership.

```{r fuzz-02, exercise=TRUE}
head(round(cl$memb, 2))               # fuzzy membership
(grp <- defuzzify(cl, 'cut', alpha=0.8)[[2]]) # crisp, at threshold (incl NAs)
(grp <- defuzzify(cl, 'max')[[2]])    # crisp membership, at max membership
table(grp, useNA='always')            # tally points per group
```

Also examine the group characteristics:

```{r fuzz-03, exercise=TRUE}
### examine group characteristics
cl$size                               # group size (sum of fuzzy memberships)
cl$withinss                           # group sums-of-squares
round(ctr <- clustcentroid(D, grp),3) # group centroids
(medoids  <- clustmedoid(spe, grp))   # *indices* of group medoids
clustvar(D, grp)                      # within-group variance
clustvar(D)                           # pooled among-group variance
as.dist(D_ctr <- as.matrix(interclustdist(D, grp))) # dissim among centroids
as.dist(D_med <- as.matrix(D)[medoids,medoids])     # dissim among medoids
```

How are species represented per group? Examine species frequency (constancy) per group:

```{r fuzz-04, exercise=TRUE}
con <- clustconst(spe, memb=as.memb(grp))
tabasco(t(con), col=get_palette())    # constancy per group
summary(con, mode='all')              # constancy per group
summary(con, mode='cluster', 'M1')    # examine one particular group
```

How to choose a valid number of groups? Systematically vary _k_, then evaluate. PCN = normalized partition coefficient (higher is better). PEN = normalized partition entropy (lower is better).

```{r fuzz-05, exercise=TRUE}
### systematically vary k from 3 to 9, then evaluate
cl_k <- random.vegclustdist(D, cmin=3, cmax=9, nstart=3, method='FCM', m=1.2)
sapply(seq_along(cl_k), function(i) vegclustIndex(cl_k[[i]])) # evaluate!
```

You may wish to add new sites to an existing classification.  For example, imagine we fit a model using "western" sites, and now must assign values for the "eastern" sites.

```{r fuzz-06, exercise=TRUE, fig.asp=0.667}
### assign *new* sites to an *existing* classification ---
i    <- (xy$x < 220)                  # split the dataset, based on location
west <- spe[i,]                       # old calibration data = west
west <- west[,colSums(west) > 0]      # exclude zero-sum species
east <- spe[!i,]                      # new evaluation data = east
east <- east[,colSums(east) > 0]      # exclude zero-sum species
k    <- 7                             # number of groups specified in advance
(grp_west <- cutree(hclust(vegdist(west), 'ward.D2'), k)) # *existing* model
cl_west   <- as.vegclust(vegdist(west), grp_west) # convert to vegclust object
m  <- conformveg(west, east)                      # merge datasets
DD <- as.matrix(vegdist(rbind(m$x,m$y)))          # ALL dissimilarities
DD <- DD[(NROW(west)+1):NCOL(DD), 1:NROW(west)]   # rows = new, cols = old
cl_east <- vegclass(cl_west, DD)      # assign new points given existing model
grp_west <- defuzzify(cl_west)[[2]]   # memberships of *old* points
grp_east <- defuzzify(cl_east)[[2]]   # memberships of *new* points
grp <- c(grp_west,grp_east)
par(mfrow=c(1,2), oma=c(0,0,0,0), mar=c(2,2,1,0))
plot(xy, col=i+1, pch=16)             # map east/west locations
plot(xy, col=colvec(grp), pch=16)     # map group memberships
text(xy, grp)                         # identify group membership
legend('bottomleft', leg=1:k, fill=unique(colvec(grp)), title='Group', cex=0.7)
```

### Optimal partitioning
```r 
# optimal partitioning
require(optpart)

```


### Cluster evaluation and goodness-of-fit


### Key references

de Caceres, M., X. Font, F. Oliva. 2010. The management of numerical vegetation classifications with fuzzy clustering methods. _Journal of Vegetation Science_ 21: 1138–1151.

Murtagh, F., and P. Legendre. 2014. Ward's hierarchical agglomerative clustering method: which algorithms implement Ward's criterion? _Journal of Classification_ 31: 274–295.


<!-- --------------------------------------------- -->


## Group differences

#### __*Do community compositions significantly differ among groups?*__ {#permanova}

Let's assume PERMANOVA is the best choice for comparing groups in multivariate space. Other methods (ANOSIM, MRPP) exist, but PERMANOVA usually has better power and false-detection rates, even when groups may have different dispersions (Anderson and Walsh 2013) among other operational advantages.

```{r grp-define, exercise=TRUE}
### define and visualize four groups
brk     <- quantile(env$k, probs=seq(0,1,by=0.25))   # define breaks
grp_k <- cut(env$k, brk, include.lowest=T,
               labels=c('lo','med','hi','veryhi'))   # group memberships
table(grp_k, useNA='always')                       # group tally
tapply(env$k, INDEX = grp_k, FUN = mean)           # group means
plot(m1$points, pch=NA)                              # visualize on the NMS
text(m1$points, labels=grp_k, col=as.numeric(grp_k)) # group memberships
ordispider(m1, groups=grp_k, col=1:4)              # group centroids 
```

### Test for difference in community compositions

PERMANOVA tests for differences in multivariate centroid, _in the space of the chosen dissimilarity measure_.  Recall that matrix `D` was our matrix of pairwise Bray-Curtis dissimilarities.

```{r grp-diff-permanova, exercise=TRUE}
### permanova: test for differences in multivariate *centroid*
a1 <- adonis(D ~ grp_k, permu=999)
a1
```

How do we interpret this result?  Based on Bray-Curtis dissimilarities, four groups defined by soil potassium significantly differed in community compositions (PERMANOVA pseudo-_F_ = 10.2, permutational _p_ = 0.001, _R_^2^ = 0.25).  The NMS ordination helps us see that this is largely driven by the "veryhi" group.  Which you can test formally by repeating PERMANOVA for pairwise multiple comparisons.

### Test for homogeneity of community compositions

A related test, PERMDISP, tests for differences in multivariate dispersion, _in the space of the chosen dissimilarity measure_. This is interpreted as a test of among-group homogeneity of community compositions, i.e., beta-diversity.  Conveniently, we can ask for pairwise multiple comparisons.

```{r grp-diff-permdisp, exercise=TRUE}
### permdisp: test for differences in multivariate *dispersion*
b1 <- betadisper(D, grp_k)
b1
permutest(b1, pairwise=TRUE, permu=999)
boxplot(b1)
```

How do we interpret this?  Based on Principal Coordinates reduction of original Bray-Curtis dissimilarities, the four groups defined by soil potassium significantly differed in homogeneity of community compositions (PERMDISP _F_ = 7.98, permutational _p_ = 0.001).  This was due to the "veryhi" potassium group having significantly lower dispersion (lower beta-diversity) than the lower-potassium groups.

### PERMANOVA equivalence to ANOVA

PERMANOVA works as 'regular' ANOVA when using Euclidean distances.  We'd expect a different _p_-value since we're using permutations, but identical _F_-values and sums-of-squares.

```{r grp-diff-equivalence, exercise=TRUE}
De <- dist(env$k, 'euc')            # Euclidean distances
adonis(De ~ grp_k, perm=999)      # examine F and SS
print(anova(lm(k ~ grp, env)))      # expect identical F and SS
```

### PERMANOVA for blocked design

For blocked designs (e.g., for repeated measures or genuine blocked designs), we must specify how permutations occur.  Because units are only exchangeable _within blocks_, the permutations should occur within blocks.

```{r grp-diff-blocked, exercise=TRUE}
### blocked design: permutations must occur within strata
blk <- factor(LETTERS[sample(rep(1:12,len=nrow(env)))]) # arbitrary 'blocks'
plot(m1)                                       # visualize on NMS
ordispider(m1, blk, label=TRUE)
customperm <- how(nperm=999)                   # set number of permutations
setBlocks(customperm) <- blk                   # permute only *within* blocks
adonis(D ~ grp_k, permutations=customperm)   # correct test
```

### Key references

Anderson, M. J., and D. C. I. Walsh. 2013. PERMANOVA, ANOSIM, and the Mantel test in the face of heterogeneous dispersions: What null hypothesis are you testing? *Ecological Monographs* 83:557–574. https://doi.org/10.1890/12-2010.1


<!-- --------------------------------------------- -->


## Group indicator species

#### __*Which species are indicative of pre-defined groups?*__ {#indval}

The Indicator Value (_IndVal_) is the product of a species' fidelity and specificity to a group (Dufrêne and Legendre 1997).

Here we use `labdsv::indval` to examine indicators of _single_ groups, but also see R package `indicspecies` to test whether species might indicate _multiple_ groups.

First, let's use the same groups we used for PERMANOVA: which species are indicative of the four groups defined by soil potassium?  We can compare these "real" indicator species against a random expectation.

```{r indval, exercise=TRUE}
### real groups
grp <- cut(env$k, quantile(env$k, probs=seq(0,1,by=0.25)), include.lowest=T,
               labels=c('lo','med','hi','veryhi'))   # group memberships
iv  <- labdsv::indval(spe, grp) # indicator species analysis for *real* groups
summary(iv)                     # IndVal observed

### random groups
rnd <- sample(grp, length(grp), replace=T) # define random groups by bootstrapping
ivr <- labdsv::indval(spe, rnd) # indicator species analysis for *random* groups
summary(ivr)                    # IndVal expected at random
```

How many indicator species would you expect at random?

```{r indval-null}
### null expectation setting alpha = 5%
ceiling( ncol(spe) * 0.05 )

```

If you defined groups by a clustering method using species information, do not use _IndVal_ to find indicator species of those groups (that would be circular logic).  


### Key references

Dufrêne, M. and Legendre, P. 1997. Species assemblages and indicator species: the need for a flexible asymmetrical approach. _Ecological Monographs_ 67(3): 345--366.  

de Cáceres, M., P. Legendre, and M. Moretti. 2010. Improving indicator species analysis by combining groups of sites. _Oikos_ 119:1674–1684.


<!-- --------------------------------------------- -->


## Community traits

#### __*How do species’ evolutionary relationships explain responses?*__ {#traits-main}

Recall the traits matrix, which assigns each species (rows) a value for each of twelve traits (columns):

```{r peek-traits, exercise=T}
names(tra)
head(tra)
hist(tra$lfp, breaks=11)
```

We can ask several questions about trait diversity, convergence/divergence, and relationships to environment.

### Trait dissimilarity

#### __*Are traits (for species in a site) convergent or divergent?*__ {#trait-convergent}

The mean pairwise distance (MPD) shows how far (on average) the traits of one community are relative to all others.  Most commonly these are Euclidean or Gower distances.  A randomization test gives a test statistic called SES, standardized effect size.  Significant positive SES indicates traits are less similar than expected at random (i.e., species have divergent traits), while negative SES indicates traits are more similar than expected at random (i.e., species tend to converge on similar traits). We can examine trait divergence/convergence along a gradient, or test for divergence/convergence within groups.

```{r trait-converge, exercise=TRUE}
### Euclidean trait distance (traits already scaled 0-1)
Dt  <- dist(tra, method='euc')
### calculate trait SES of mean pairwise distances in sites
ses <- picante::ses.mpd(spe, Dt, null.model='richness', abund=F, runs=999)
### plot SES across the gradient
plot(ses$mpd.obs.z ~ env$k, ylab='Trait SES(MPD)', xlab='Soil potassium', 
     col=ifelse(ses$mpd.obs.p < 0.05, 'red','black'))
abline(lm(ses$mpd.obs.z ~ env$k)) # regression line
abline(h=0, lty=2)                # random-traits line
text(0.9, 1, 'Divergent')      
text(0.9, -4, 'Convergent')
```

### Nonlinear trait-environment correlations

#### __*How strongly are traits related to environment?*__ {#trait-env-corr}

To evaluate the strength of relationship between traits and environment, we can perform ordination of sites based on traits (rather than species).  We first need to make a *community-weighted means* (*CWM*) matrix, a traits-by-site matrix whose values are trait means across the species in each site.

```{r cwm, exercise=TRUE}
`makecwm` <- function (spe, tra) {
  spe <- as.matrix(spe)
  tra <- as.matrix(tra)
  `standardize` <- function(x) {
    (x - min(x, na.rm=TRUE)) / diff(range(x, na.rm=TRUE))
  }
  tra <- apply(tra, MARGIN = 2, FUN = standardize)
  awt <- spe %*% tra                 # abundance-weighted trait totals
  awt / rowSums(spe, na.rm=TRUE)     # community-weighted traits matrix
}
cwm <- data.frame(makecwm(spe, tra)) # make the CWM traits matrix
tabasco(cwm, col=get_palette())      # visualize
```

Ordination of the CWM matrix gives a configuration of sites in *traits space*, where distance between points indicates dissimilarity of trait compositions (trait syndromes).  Nonlinear regression of environmental variables in this traits space can give us the trait-environment correlations that we're after.

```{r trait-ordination, exercise=TRUE}
### NMS ordination of traits
m <- metaMDS(cwm, 'altGower', k=2, trace=0)
plot(m, cex=cwm$lfp) # plot sites in *traits space*

### fit one environmental variable in traits space
o <- ordisurf(m, env$k, col=3, add=T)

### fit *all* environmental variables in traits space
fit <- t(sapply(env, function(i) {
  g <- summary(vegan::ordisurf(m ~ i, plot=F)) # fit GAM
  c(`pval` = as.numeric(sprintf('%.3f',round(g$s.pv,3))),
    `r2`   = as.numeric(sprintf('%.3f',round(g$r.sq,3))))
}))
cat('GAM p-values and goodness-of-fit, sorted\n',
    '----------------------------------------\n')
fit[rev(order(fit[,2])),]
```

Here, the environmental variables with the strongest _nonlinear_ relationship to trait compositions were potassium, sodium, and elevation.  

### Fourth-corner analysis

RLQ and fourth-corner analysis can also detect linear traits-environment relationships as mediated by the species abundance matrix.  There is no need to calculate the CWM matrix.  Because it boils down to a series of PCA operations, the RLQ method does not readily admit nonlinear relationships.  It is included here for completeness.

```
### the RLQ method
o_spe <- dudi.coa(spe, scannf = FALSE, nf = 2)
o_env <- dudi.hillsmith(env, scannf = FALSE, nf = 2, row.w = o_spe$lw)
o_tra <- dudi.hillsmith(tra, scannf = FALSE, nf = 2, row.w = o_spe$cw)
r     <- rlq(o_env, o_spe, o_tra, scannf = FALSE, nf = 2)
plot(r)
summary(r)
randtest(r)
fourthcorner.rlq(r, type='Q.axes')
fourthcorner.rlq(r, type='R.axes')
```

### Phylogenetic correction of traits

#### __*How to remove phylogenetic relatedness from traits?*__ {#trait-phylo-correct}

Often we suspect that close phylogenetic relatives may share similar trait values than more distant relatives (i.e., trait conservatism).  For further analyses it may be useful to "remove" the effects of relatedness from traits, based on the phylogenetic variance-covariance matrix.

```{r phylo-correct}
### function to do phylogenetic correction
`phylo_corr` <- function(phy, tra, ...){
  if (class(phy) != 'phylo') stop('phy must be of class `phylo`')
  if (!is.data.frame(tra)) tra <- as.data.frame(tra)
  rn <- dimnames(tra)[[1]] # species names
  cn <- dimnames(tra)[[2]] # traits names
  n  <- dim(tra)[[2]]    # n of traits
  if (!identical(phy$tip.label, rn)) stop('species name mismatch')
  G  <- ape::vcv.phylo(phy) # phylo variance-covariance matrix
  corfac <- chol(solve(G)) # 'correction factor' per Butler et al. (2000) 
  U  <- as.data.frame(      # initialize U for 'corrected' trait values
    matrix(NA, nrow=dim(tra)[[1]], ncol=dim(tra)[[2]],
           dimnames=list(rn,cn)))
  tra <- droplevels(tra)  # drop unused factor levels
  tra[,sapply(tra,function(x)nlevels(x)==1)] <- 1 # correct 1-level factors
  for(j in 1:n){       # corrections per individual trait
    M       <- model.matrix(as.formula(paste0("~0+",cn[j])), data=tra)
    corrtra <- data.frame(corfac %*% M)
    U[,j]   <- corrtra[rn, ,drop=FALSE]
  }
  return(U)
}
### do the correction, then compare
p <- phylo_corr(phy, tra) 
par(mfrow=c(2,2))
for(i in 7:10) {
  plot(tra[,i], p[,i], pch=16, cex=0.7, col='#00000050',
       main=dimnames(p)[[2]][i], xlab='Original', ylab='Corrected')
}
```


### SYNCSA

#### __*How do traits, environment and phylogenies relate to community assembly?*__ {#syncsa-method}

The SYNCSA approach (Pillar and Duarte 2010) uses matrix correlations among species abundances, traits, phylogenetic relatedness, and environment as evidence of different community assembly processes.  As in all of science, recall that correlation is not causation.

```{r syncsa, exercise=TRUE}
# subset traits, for convenience
tra <- tra[,c('annual','biennial','perennial','lfp','max_height','bfp')] 
# run SYNCSA
res <- syncsa(comm       = spe,
               traits    = tra,
               phylo     = cophenetic(phy),
               envir     = env,
               ro.method = 'procrustes',
               permu     = 199)
# view result
res
```

### Key references

Butler, M.A., T.W. Schoener, and J.B. Losos. 2000. The relationship between sexual size dimorphism and habitat use in Greater Antillean _Anolis_ lizards. *Evolution* 54:259-272.  

Eklöf, A., and D.B. Stouffer. 2016. The phylogenetic component of food web structure and intervality. *Theoretical Ecology* 9:107–115.

Pillar, V.D., and L.D.S. Duarte. 2010. A framework for metacommunity analysis of phylogenetic structure. *Ecology Letters* 13:587–596.


<!-- --------------------------------------------- -->


## Community phylogenetics

#### __*How do species' evolutionary relationships explain responses?*__ {#phylo-main}

### Plot phylogenies

```{r phylo-plot, exercise=TRUE}
### basic plotting
par(mfrow=c(1,3), mar=c(0,0,0,0), oma=c(0,0,0,0)) # plotting parameters
phy                                               # basic structure
plot(phy, cex=0.75, no.margin=T)                  # basic plotting

### color tip labels by trait values
plot(phy, cex=0.75, tip.color=colvec(tra$lfp), no.margin=T)
# axisPhylo(cex=0.6)

### symbolize trait values at tips
plot(phy, cex=0.75, label.offset = 48, no.margin=T)
tiplabels(pch = 21, bg = c(tra$annual), adj = 0)
tiplabels(pch = 21, bg = c(tra$biennial), adj = 15)
tiplabels(pch = 21, bg = c(tra$perennial), adj = 30)
```


### Phylogenetic diversity - alpha

According to some people, phylogenetic diversity lets you infer community assembly processes (competitive exclusion, environmental filtering, neutral dynamics).  Some other serious people doubt this.

```{r phy-div-alpha, exercise=T}
### phylogenetic diversity metrics
Dp  <- cophenetic(phy) # phylogenetic distances
### Faith's PD phylogenetic diversity = total branch length connecting all species in a site
fpd <- pd(spe, phy)    # Faith's PD
### mean pairwise distance
mpd <- ses.mpd(spe,  Dp, null.model='independentswap') 
### mean nearest taxon distance
mnd <- ses.mntd(spe, Dp, null.model='independentswap') 
### bring all together
phy_div <- cbind(fpd, mpd = mpd$mpd.obs.z, mntd = mnd$mntd.obs.z)
head(phy_div)
pairs(phy_div)
```


### Phylogenetic diversity - beta

Phylogenetic Bray-Curtis dissimilarities = fraction of branch-length shared between communities.  Different from the cophenetic branch lengths (between-species), this shows sample unit dissimilarities (between-sites).

```{r phy-div-beta, exercise=T}
### correlation between phylogenetic and taxonomic beta-diversity
Dp <- phylosor(spe, phy) # phylogenetic distances
protest(Dp, D)           # procrustes correlation

### side-by-side NMS ordinations
par(mfrow=c(1,2))
plot(scores(metaMDS(D)), col=colvec(1:97), pch=19, main='Taxonomic NMS')
plot(scores(metaMDS(Dp)), col=colvec(1:97), pch=19, main='Phylogenetic NMS')
```

### Phylogenetic signal

#### __*How similar are traits among related species?*__ {#phylo-signal}

We focus on a measure called Blomberg's _K_ to detect phylogenetic signal in traits (Revell et al. 2008).  A value of _K_ > 1.0 indicates that traits of related species are more similar than expected from Brownian motion, and _K_ < 1.0 indicates traits of related species are more dissimilar than expected at random.

```{r phylo-signal, exercise=TRUE}
### detect phylogenetic signal for traits
K <- sapply(tra, FUN=function(j){
  names(j) <- rownames(tra)
  round(picante::phylosignal(j, ape::multi2di(phy)),6)})
as.matrix(K)
### Thinker: which traits have strong evidence for conservatism?
```

### Key references

Revell, L.J., L.J. Harmon, and D.C. Collar. 2008. Phylogenetic signal, evolutionary process, and rate. _Systematic Biology_ 57:591–601.


<!-- --------------------------------------------- -->


## Community spatial analysis

#### __*To what degree are communities structured spatially?*__ {#spatial-mantel}

### Mantel test

The Mantel test (Mantel 1967) compares a community dissimilarity matrix to a spatial distance matrix (it can also be used to compare any two dissimilarity matrices, for example, to test goodness-of-fit of Euclidean ordination distances to original community dissimilarities, or two ordination configurations).  Here we ask, is compositional dissimilarity related to spatial distances?

```{r mantel-orig, exercise=TRUE}
E <- dist(xy)                            # spatial distance matrix
vegan::mantel(D, E, method='spearman')   # spearman *rank* correlation
# in `ecodist` package, we also get useful bootstrap CIs:
ecodist::mantel(D ~ E, mrank=T)          # spearman *rank* correlation
```

A Mantel correlogram depicts this correlation of compositional dissimilarity at specified lag distances.

```{r mantel-correlogram, exercise=TRUE}
brk <- seq(0, round(max(E),-1), by=10)
plot(ecodist::mgram(D, E, breaks=brk, nperm=99, mrank=T, nboot=500))
abline(h=0, col='red')
plot(vegan::mantel.correlog(D, E, break.pts=brk, cutoff=F, r.type='spearman',
                            nperm=99, mult='holm', progressive=T))
```

### MRM: multiple regression on distance matrices

MRM: multiple regression on distance matrices (Lichstein 2007).

```{r mrm, exercise=TRUE}
# species dissimilarities are related to space (extremely weakly) and 
#   potassium (moderately)
ecodist::MRM(D ~ dist(xy) + dist(env$k), nperm=999) 

# abundance of cosmopolitan bulrush is NOT related to space, 
#   but is related to potassium (moderately)
ecodist::MRM(dist(spe$bolboschoenus_maritimus) ~ dist(xy) + dist(env$k), nperm=999) 

```


### PCNM: principle coordinates of neighbor matrices.

#### __*How much variation is explained by space, environment, or spatially-structured environment?*__ {#spatial-pcnm}

We can quantify spatial distances among sites with a Euclidean distance matrix. Performing principal coordinates analysis on this "neighbor matrix" gives us PCNMs: principle coordinates of neighbor matrices.  PCNM scores are interpreted as *orthogonal* explanatory variables at varying spatial scales: think of them as progressively broad-, medium- and local-scale predictors.

```{r pcnm-mapping, exercise=TRUE, fig.asp=0.555}
E   <- dist(xy)            # euclidean distances between sites
pc  <- pcnm(E)             # principal coordinates of neighbor matrices
par(mfrow=c(2,3), oma=c(0,0,0,0), mar=c(2,2,2,0)) # map PCNMs across study area
v   <- 2^(0:5)             # select PCNMs at increasingly fine scale
for(i in seq_along(v)) {
  ordisurf(xy, scores(pc, choices=v[i]), bubble=3, main=paste0('PCNM ',v[i]))
}
```

We can use *weighted* principle coordinates of neighbor matrices as explanatory variables in CCA to construct a multiscale ordination (MSO, Wagner 2004) variogram.  Residual variance here would indicate spatial autocorrelation; we hope for this to exhibit no trend with distance, since PCNMs account for spatial autocorrelation.

```{r pcnm-mso, exercise=TRUE}
rs    <- rowSums(spe) / sum(spe)  # sites weighted by abundances 
pc    <- pcnm(E, w=rs)            # *weighted* PCNMs
ord   <- cca(spe ~ scores(pc))    # CCA: species constrained by space
msord <- mso(ord, xy)             # multiscale ordination
msoplot(msord, expl=T)            # MSO variogram: spatial trend in residuals?
```

Using the weighted PCNMs above, we can perform variance partitioning of species dissimilarities in dbRDA.  This particular variation partitioning shows the amount of variation in species dissimilarities (**D**) explained by 1) space, 2) environment, and 3) spatially-structured environment.

```{r pcnm-varpart-01, exercise=TRUE}
(vp <- varpart(D, env, scores(pc))) 
plot(vp, bg = c(2,4))
text(-0.2, 0.3, 'Environment', cex=0.7)
text(0.5, 0.3, 'Spatially-\nstructured\nenviro', cex=0.7)
text(1.2, 0.3, 'Space', cex=0.7)
```

We can include more explanatory matrices -- here, species dissimilarities explained by 1) environment, 2) broad spatial, and 3) local spatial.

```{r pcnm-varpart-02, exercise=TRUE}
pc_broad <- scores(pc)[,1:10]
pc_local <- scores(pc)[,11:59]
(vp <- varpart(D, env, pc_broad, pc_local)) 
plot(vp, cex=0.7, bg=2:5) 
text(-0.2, 0.3, 'Environment', cex=0.7)
text(0.5, 0.55, 'Broad-\nstructured\nenviro', cex=0.7)
text(1.2, 0.3, 'Broad\nspatial', cex=0.7)
text(-0.5, -0.75, 'Local-\nstructured\nenviro', cex=0.7)
text(0.5, -1.15, 'Local\nspatial', cex=0.7)
### Thinker: why do "broad" and "local" share in common zero variance explained?
```

We can include more explanatory matrices -- here, species dissimilarities explained by 1) soil fractions, 2) soil chemistry, and 3) spatial processes.

```{r pcnm-varpart-03, exercise=TRUE}
(vp <- varpart(D,                               # dissimilarities
               ~ clay + sand + silt,            # soil fractions
               ~ mg + k + conductivity + na_l,  # soil chemistry
               scores(pc),                      # spatial predictors
               data=env))
plot(vp, cex=0.7, bg=2:5)
text(-0.2, 0.3, 'Soil fractions', cex=0.7)
text(1.2, 0.3, 'Soil chemistry', cex=0.7)
text(0.5, -1.15, 'Spatial', cex=0.7)
text(-0.5, -0.75, 'Spatially-\nstructured\nsoil fractions', cex=0.7)
text(1.5, -0.75, 'Spatially-\nstructured\nsoil chemistry', cex=0.7)
### Thinker: how do you interpret this figure?  
### Moderate spatial structuring, a bit owing to soil chemistry,
### but mostly to other unmeasured spatial processes (dispersal?).

### Thinker: what does this NOT explain?  
### What is the biological cause of residual variation?
### We did ordination in 1) species space, 2) env space, 
###   3) env constrained spp space?
```

### Key references

Borcard D. and Legendre P. 2002. All-scale spatial analysis of ecological data by means of principal coordinates of neighbour matrices. _Ecological Modelling_ 153: 51-68.

Dray, S., R. Pélissier, P. Couteron, M.-J. Fortin, P. Legendre, P. R. Peres-Neto, E. Bellier, R. Bivand, F. G. Blanchet, M. De Cáceres, A.-B. Dufour, E. Heegaard, T. Jombart, F. Munoz, J. Oksanen, J. Thioulouse, and H. H. Wagner. 2012. Community ecology in the age of multivariate multiscale spatial analysis. _Ecological Monographs_ 82: 257-275.

Lichstein, J. 2007. Multiple regression on distance matrices: a multivariate spatial analysis tool. _Plant Ecology_ 188:117-131.

Mantel, N. 1967. The detection of disease clustering and a generalized regression approach. _Cancer Research_ 27:209-220.

Wagner, H.H. 2004. Direct multi-scale ordination with canonical correspondence analysis. _Ecology_ 85: 342-351.


<!-- --------------------------------------------- -->


## Appendix 1: Workflow example {#workflow}

This workflow uses common best-practices in ecological community analysis.  It is somewhat opinionated, and is by no means universal or obligatory!

**Data I/O:** read data into R (typically `read.csv` if CSV files, or `load` if in R binary format).

**Basic exploration:** use basic commands to examine data structure (`str`), peek at data (`head`), see the frequency distribution (`hist`), or make biplots (`plot` or `pairs`).

**Advanced exploration:** `vegan::tabasco` or `labdsv::abuocc` to visualize species abundance matrices.

**Transformations:** use `vegan::tabasco` for relativizations, and arithmetic for transformations.  Try log~10~(_x_+1) for abundances, row-standardize abundances to put sites on equal footing, and range-standardize for environment or traits (so all on same scale) -- McCune and Grace 2002

**Outliers:** 
sample units with mean dissimilarities > defined number of SDs from grand mean -- McCune and Grace 2002

**Community dissimilarities:** use `vegan::vegdist` to get Bray-Curtis dissimilarities -- Legendre and Legendre 2012

**Environmental distances:** use `scale` and `dist` to get (scaled) Euclidean distances

**Dissimilarity adjustments:** use `vegan::stepacross` for step-across shortest distance in the presence of high beta-diversity -- Smith 2017

**Ordination of sites, based on communities:** use `vegan::metaMDS` for NMS -- Minchin 1987

**Ordination of sites, based on environment:** use `stats::prcomp` for PCA -- linear combinations

**Ordination of sites, based on communities constrained by environment:** use `vegan::dbrda` for dbRDA, or `fso::mfso` for FSO -- McArdle and Anderson 2001 or Roberts 2008

**Comparing ordination configurations:** use `vegan::protest` for Procrustes analysis -- Peres-Neto and Jackson 2001

**Clustering:** use `vegclust::vegclust` for fuzzy clustering, or `optpart::optpart` for optimal partitioning -- de Cáceres 2010 or Roberts 2015

**Group differences:** use `vegan::adonis2` for PERMANOVA -- Anderson and Walsh 2013

**Group indicator species:** use `labdsv::indval` for the usual single-group _IndVal_, or `indicspecies::multipatt` for multi-group _IndVal_ -- Dufrene and Legendre 1997 or 
de Cáceres et al. 2010

<!-- --------------------------------------------- -->


## Appendix 2: Further resources

#### General R usage

[R for Data Science](https://r4ds.had.co.nz/)  

[Data Carpentry](https://datacarpentry.org/R-ecology-lesson/index.html)  

#### Ecology in R

[Dave Roberts' labdsv pages](http://ecology.msu.montana.edu/labdsv/R/)  

[Jari Oksanen's vegan introduction - PDF](https://cran.r-project.org/web/packages/vegan/vignettes/intro-vegan.pdf)  

[David Zelený's community analysis](http://www.davidzeleny.net/anadat-r/doku.php?id=en:start)  

[Tad Fukami's ecological statistics](https://fukamilab.github.io/BIO202/)  

[Steve Kembel traits and phylogenies tutorial](http://www.kembellab.ca/r-workshop/biodivR/SK_Biodiversity_R.html)

#### Community analysis R packages

[CRAN Environmetrics Task View](https://cran.r-project.org/web/views/Environmetrics.html)  
[vegan](https://cran.r-project.org/web/packages/vegan/index.html)  
[labdsv](https://cran.r-project.org/web/packages/labdsv/index.html)  
[optpart](https://cran.r-project.org/web/packages/optpart/index.html)  
[ade4](https://cran.r-project.org/web/packages/ade4/index.html)  
[ecodist](https://cran.r-project.org/web/packages/ecodist/index.html)  


<!-- --------------------------------------------- -->


<!-- ## Appendix 3: FAQ -->

<!-- [Why are we here today?](#why-here) -->

<!-- [Where did this dataset originate? How do I use it?](#dataorigin) -->

<!-- [Where did this dataset originate?](#data-origin) -->

<!-- [How do I get my data into shape for analysis?](#pre-analysis) -->

<!-- [How diverse is this site? many sites? region?](#diversity-main) -->

<!-- [How dissimilar are community compositions among sites?](#dissim-main) -->

<!-- [How best to order sites by dissimilarities of community compositions?](#ord-unconstrained) -->

<!-- [How best to order sites by dissimilarities of community compositions, in response to measured environmental variables?](#ord-constrained) -->

<!-- [How to group sites according to dissimilarity of community compositions?](#clustering) -->

<!-- [Do community compositions significantly differ among groups?](#permanova) -->

<!-- [Which species are indicative of pre-defined groups?](#indval) -->

<!-- [Are traits (for species in a site) convergent or divergent?](#trait-convergent) -->

<!-- [How strongly are traits related to environment?](#trait-env-corr) -->

<!-- [How to remove phylogenetic relatedness from traits?](#trait-phylo-correct) -->

<!-- [How do traits, environment and phylogenies relate to community assembly?](#syncsa-method) -->

<!-- [How do species' evolutionary relationships explain responses?](#phylo-main) -->

<!-- [How similar are traits among related species?](#phylo-signal) -->

<!-- [To what degree are communities structured spatially?](#spatial-mantel) -->

<!-- [How much variation is explained by space, environment, or spatially-structured environment?](#spatial-pcnm) -->

<!-- [What is a typical workflow for ecological community analysis in R?](#workflow) -->
















<!-- --------------------------------------------- -->

<!-- ## Direct gradient analysis -->

<!-- #### *How does this species (or trait, or ...) relate to an environmental gradient?* -->

<!-- ```{r direct-gradient-analysis, exercise=TRUE} -->
<!-- # prepare vector data -->
<!-- y <- rowSums((spe > 0) * 1)  # response vector = species richness -->
<!-- x <- env[,'k']               # predictor vector = potassium concentration -->

<!-- # fit two models -->
<!-- mod_linear    <- stats::lm(y ~ x)                # simple linear model -->
<!-- mod_nonlinear <- mgcv::gam(y ~ s(x), gamma=0.25) # nonlinear GAM model -->

<!-- # view model summaries -->
<!-- summary(mod_linear) -->
<!-- summary(mod_nonlinear) -->

<!-- # visualize fitted values -->
<!-- plot(x, y, xlab='Potassium', ylab='Richness', pch=16, col='#00000050') -->
<!-- xs <- seq(min(x), max(x), len = 99) -->
<!-- lines(xs, predict(mod_linear, data.frame(x = xs)), lwd = 2, col='darkblue') -->
<!-- lines(xs, predict(mod_nonlinear, data.frame(x = xs)), lwd = 2, col='orange') -->

<!-- # LETS PREDICT onto a landscape grid (raster) -->
<!-- # first, simulate data with spatial pattern (negative exponential decay) -->
<!-- `simulate_grid` <- function(nr = 49, seed = 2021, ...) { -->
<!--   g <- expand.grid(lon=seq(0,1,len=nr),lat=seq(0,1,len=nr)) # all combos -->
<!--   `f` <- function(n, mu = 0, V = matrix(1)) { # multivariate normal sampling -->
<!--     p <- length(mu) -->
<!--     t(matrix(rnorm(n * p), ncol=p) %*% chol(V) + rep(mu, rep(n, p))) -->
<!--   } -->
<!--   set.seed(seed=seed) -->
<!--   k <- f(1, rep(3,nrow(g)), exp(-0.05 * as.matrix(dist(g)))) # do sampling -->
<!--   k <- decostand(scale(k, center=F), 'range')                # re-scale 0-1 -->
<!--   return(cbind(g,k)) # soil value at each lon/lat -->
<!-- } -->
<!-- nr <- 49 -->
<!-- g <- simulate_grid(nr) -->

<!-- # predict from the models -->
<!-- g$pr_lin <- predict(mod_linear, newdata=data.frame(x=g$k)) -->
<!-- g$pr_non <- predict(mod_nonlinear, newdata=data.frame(x=g$k)) -->

<!-- # plot the landscapes -->
<!-- u  <- c('#414487','#39558C','#30698E','#297A8E','#228B8D','#1E9C89', -->
<!--         '#26AD81','#40BD72','#65CB5E','#90D743','#BBDF27') # colors -->
<!-- par(mfrow=c(1,3)) -->
<!-- image(matrix(g$k, nrow=nr), col=u, main='patchy landscape\nof potassium') -->
<!-- image(matrix(g$pr_lin, nrow=nr), col=u, main='species richness,\nlinear') -->
<!-- image(matrix(g$pr_non, nrow=nr), col=u, main='species richness,\nnonlinear') -->
<!-- ``` -->






<!-- ### Best subset of environmental predictors -->

<!-- Which environmental predictors are most related to community compositions?  We can find this by maximizing the rank correlation between environmental distances and community dissimilarities, using the BIOENV procedure. -->

<!-- ```{r best-bioenv, exercise=TRUE} -->
<!-- best <- bioenv(D, env=env[,sapply(env, is.numeric)], upto=3, trace=T) -->
<!-- best -->
<!-- ``` -->



